Data Management \\ and LSST Special Programs

### 2022-02-20 DMTN-065.tex
doc rev req

\setDocAbstract{
The core LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, which is expected to be accomplished with 85--90\% of the observing time.
The remaining 10--15\% of the time will be spent on Special Programs: additional survey areas and/or observing strategies that are driven by specific science goals which either build on, or are outside of, the core science pillars of the LSST.

This document provides a summary of the Special Programs being considered, the potential diversity of data they might produce (e.g., shorter or longer exposure times), and the role of Rubin Observatory in reducing, processing, and serving data products from Special Programs (including the flow-down of related requirements, and the capabilities of the LSST Science Pipelines and the Rubin Science Platform).

The specialized, science-specific aspects of Special Programs processing and analysis that are best left to the science community are also described, and illustrated with case-study examples.

The main target audience of this document is Rubin Observatory staff -- the construction-era Data Management (DM) team and the operations-era Data Production (RDP) team -- but members of the science community who are planning to use Special Programs to reach their science goals may also find this document useful.
}

@@

Note that there remains some grey area between what is a mini-survey within the WFD, and what is a WFD-like micro-survey.
For example, the last two on the list above (the northern stripe and the Virgo cluster) could also (or might be in the future) considered WFD mini-surveys and not Special Programs.

@@

\subsection{Requirements Related to Special Programs}

{\it Use the docrev.tex file to update this.}

\citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).

\subsection{Special Programs in the DPDD}

{\it Just quote what the DPDD says. Maybe reference any relevant RFC/LCR that made those changes?}

@@

{\bf Solar System Processing -- } Since Solar System Processing takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
As discussed under "Solar System Objects (SSO)" in Appendix \ref{ssec:proc_datadiv_prev}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
%There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.

@@

\begin{itemize}
\item discoverability of SP data when browsing an all-sky map
\end{itemize}

@@

\clearpage
% % % % % % % % % % % % % % % % % %
\bibliography{local,lsst,refs,books,refs_ads}

\clearpage
% % % % % % % % % % % % % % % % % %
\appendix

% % % % % % % % % % % % % % % % % %
\section{Potential Hardware Boundaries on Data Diversity}\label{sec:hardbounds}

The potential boundaries on the diversity of data products that could be imposed by limitations from the Rubin Observatory hardware -- camera, telescope, and/or site -- are considered.

\subsection{Filter Changes}
The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
Assuming that most Special Programs would be designed to keep overheads $<$100\% and would be using standard 30 second visits, the filter change time indicates that it is likely that at least 4 exposures in a given filter would be obtained between filter changes, but this is not actually a hardware boundary. 
The filter change mechanism is designed to undergo a total of 100000 changes over its lifetime, and each filter is designed to support up to 30000 changes over its lifetime, where lifetime is 15 years.
That is an average of $\sim$27 changes per day, some of which would occur in they day during calibrations (estimate, $\sim$10) and the rest at night.
As stated in the filter change memorandum (\url{ls.st/spt-494}), {\it ``the system could support as many changes involving the 5 filters loaded in the carousel as desired, without any practical limitation"}.

\subsection{Filter Carousel Loads}
The filter carousel can hold five of the six LSST filters at a time.
The system is designed to support $3000$ loads in $15$ years (\url{ls.st/spt-494}).
Filter loads are only done in the day, and there will never be data in more than five filters in a given night.

\subsection{Exposure Times}
The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}).
The maximum exposure time is not restricted.

\subsection{Readout Time}
The readout time is $2$ seconds, and would be significant overhead on short exposures.

\subsection{Inter-Image Time}
Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ seconds for thermal tolerance; i.e., the minimum readout rate might be one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
The $15$ interval between images is a potential hardware boundary on the potential diversity of data products.

\subsection{Telescope Slew}
As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no hardware boundaries on the size of a single slew or the accrued slew distance.

\subsection{Telescope Tracking}
The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.
This capability will include angular rates of up to 220 arcseconds per second in both azimuth and elevation.

\subsection{Camera Rotation}
The requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
Currently, there are no hardware boundaries imposed by camera rotation constraints on the potential diversity of data products.

% % % % % % % % % % % % % % % % % %
\section{Potential Processing Boundaries on Data Diversity}\label{sec:procbounds}

The capability of the LSST Science Pipelines to process diverse data is explored below.

Note that processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters (e.g., the number of stars with sufficient flux for photometric calibration).
Furthermore, the processing boundaries might not be fully constrained until the final performance of the LSST Science Pipelines, as described in the Data Management Applications Design, \citeds{LDM-151}) document, is fully characterized.

{\bf Summary of the conclusions below:}\\
Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
It is currently unclear whether images with very bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, or whether user generated pipelines will be needed.
The full reduction and calibration of images obtained with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope of the LSST Science Pipelines, and will require a user generated pipeline.

\subsection{Exposure Times}\label{ssec:procbounds_expt}

Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.

\subsubsection{Short Exposures (Non-Standard Visits of $<$30 sec)}
The LSST System Requirements document states that {\it ``The LSST shall be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).

However, for exposure times there are other considerations, as changing the exposure time also affects the photometric and astrometric calibrations.
Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r$$\approx$ 13--21 magnitudes.
A template image built on 15 second exposures will saturate at $r$$\approx$15.8 mag, but this still leaves stars between 15.8--21.0 magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).

In order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale).
As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").

\begin{figure}
\begin{center}
\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin. At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center. We can see that for exposure times greater than 2 seconds, the centroid converges near its true value. \label{fig:expt}}
\end{center}
\end{figure}

% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute Alert Production.
% One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
% They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
% They also present a new software package to do mitigate the effects.
% Software packages \texttt{PhoSim} (Peterson et al. 2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} could be used to characterize the PSF stability as a function of exposure time.

\subsubsection{Long Exposures (Non-Standard Visits of $>$30 sec)}

There is no maximum exposure time specified for an LSST image.
Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
Furthermore, cosmic-ray rejection completeness might be reduced for longer exposures (unknown), which could impact the quality of a difference image and the detected sources.
Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).

\subsection{Twilight Images with a Bright Background}

Images obtained during twilight for scientific purposes are also likely to have shorter exposure times, and so the issues described in Section \ref{ssec:procbounds_expt} also apply here.
Whether or not bright-background images can (or shall) be fully processed -- reduced, calibrated, background-subtracted, and delivered with astrometric and photometric solutions -- or whether this will require a user generated pipeline, remains to be determined (see also the example in Section \ref{ssec:SPCS_Twilight}).
This may depend on the exposure time and the number of stars available in the image.

\subsection{Images Obtained with Non-Sidereal Tracking}

Non-sidereal tracking leads to images in which stars are streaked, but the moving object appears as a point source.
Full processing -- providing reduced, calibrated, background-subtracted images that are delivered with astrometric and photometric solutions -- of these images is beyond the scope of the DM pipelines as it would require the development of new algorithms, and will need to be done as a user generated pipeline.
The first steps of such a pipeline, such as Instrument Signature Removal, will probably be possible to achieve by reconfiguring the relevant DM software tasks.

%%% MLG removed the following in Feb 2022, it's no longer a concern.
% \subsection{Number of Exposures per Visit (Long Sequences of a Single Field)}

% There is no processing constraint on the number of consecutive exposures that could be obtained of a single field.
% From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.

% K.-T. Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code. This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".  Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T. things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs. The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2. To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}). Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit. \textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}

% K.-T. has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler. One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence. The processing pipeline would need to know how to divide this sequence up into visits. As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.

%%% MLG removed the following in Feb 2022, it's no longer a concern for Special Programs (crowded fields are in the WFD).
% \subsection{Images in Very Crowded Fields}

% The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes.
% The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields. These may or may not be appropriate for Special Programs data, depending on the science goals.

\clearpage
% % % % % % % % % % % % % % % % % %
\section{Previously Proposed Special Programs}\label{ssec:proc_datadiv_prev}

{\bf This section has not been updated since 2018.}

In this section we compile information about the science goals and observational methods for Special Programs that have been previously proposed or discussed in the Science Community. We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science. The main resources from which we have collected information about the Community's Special Program are: \citep{2008arXiv0805.2366I}; \citep{LPM-17}; the LSST Deep Drilling Field white papers from 2011\footnote{\url{https://project.lsst.org/content/whitepapers32012}}; presentations by Niel Brandt and Stephen Ridgway at the LSST Project and Community Workshop in August 2016\footnote{\url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} and \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}}; \citep{2013arXiv1304.3455G}; and Chapter 10 of \citep{2017arXiv170804058L}.

So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}. There are three mini-survey areas that have been discussed extensively by the Science Community: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \citep{2008arXiv0805.2366I}). In Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified, along with an \textit{incomplete} list of potential mini-surveys that have been openly discussed in the Science Community. In Section \ref{sec:SPCS}, we create detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.

\begin{table}[h]
\begin{center}
\begin{footnotesize}
\caption{Approved DDF and Incomplete List of Potential Special Programs.}
\label{tab:ddfms}
\begin{tabular}{lll}
\hline \hline
Name & Coordinates & Description  \\
\hline
DDF Elias S1    & 00:37:48, -44:00:00  & approved, cadence TBD \\
DDF XMM-LSS & 02:22:50, -04:45:00  & approved, cadence TBD  \\
DDF Extended Chandra Deep Field-South & 03:32:30, -28:06:00  & approved, cadence TBD  \\
DDF COSMOS  & 10:00:24, +02:10:55 & approved, cadence TBD  \\
\hline
North Ecliptic Spur      & & solar system objects (find and characterize) \\
Galactic Plane             & & more intensive stellar surveying \\
South Equatorial Cap  & & S/LMC and more Galactic science \\
Twilight                        & & short exposures (0.1s) for bright stars \\
Mini-Moons                     &  & finding mini-moons \\
Sweetspot                       & & 60 deg from Sun for NEOs on Earth-like orbits \\
Meter-Sized Impactors     & & detection a week before impact \\
GW Optical Counterparts & & search and recovery \\
Old Open Cluster M67      & dec +12 & compact survey above Galactic plane  \\
\hline
\end{tabular}
\end{footnotesize}
\end{center}
\end{table}

Here we consider a variety of scientific fields in turn, the Special Programs that have been discussed in that Science Community so far, and the implications of these Programs for the diversity of data and data products. Generally, the types of LSST Special Programs that are open for proposals include: (i) additional deep drilling fields; (ii) refined observing strategies for deep drilling fields; (iii) optimized survey areas for the NES, South Pole, and Galactic Plane; (iv) refined observing strategies for the NES, South Pole, and Galactic Plane; and (v) additional mini-surveys (areas and observing strategies).

\medskip
\noindent \textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al. (2008, \citep{2008arXiv0805.2366I}; Section 3.1.2) describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters, repeated every two nights for four months. Each exposure would have a $5\sigma$ limit of $r\sim24$; the nightly stack would have a limit of $r\sim26.5$; and the final deep stack of all exposures would have a limit of $r\sim28$. This description does not comment on the processing mode, but, depending on the science goals the exposures could be done as either a series of 50 non-standard visits ($1\times15$ seconds) or 25 standard visits ($2\times15$ seconds).

\medskip
\noindent \textbf{Solar System Objects (SSO) -- } Four of the mini-surveys in Table \ref{tab:ddfms} have science goals related to studies of SSO. Observations of the North Ecliptic Spur area could yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (reference: Brandt's talk). The Mini-Moons Mini-Survey aims to find and study the temporarily captured satellites of the Earth (Section 10.2, \citep{2017arXiv170804058L}). The Sweetspot Survey would use twilight fields to find NEOs in Earth-like orbits (i.e., these objects are never in opposition fields, but overhead at sunrise/sunset; Section 10.2, \citep{2017arXiv170804058L}). The Meter-Sized Impactors program would find and track meter-sized impactors $<2$ weeks before impact (Section 10.2, \citep{2017arXiv170804058L}). {\bf Summary:} most of these science goals do not seem to require non-standard visits or exposure times, with the exception of the Sweetspot survey which occurs during twilight and thus may require shorter exposures. The cadence and patterns of these mini-surveys may differ from the WFD main survey, especially when very fast-moving objects are sought. From a processing perspective, it seems that many of these science goals will be achievable by using the products of Solar System Processing, which runs on the Prompt Pipeline's \texttt{DIASource} catalogs after they are updated each night. The exception is finding faint SSOs (e.g., Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) through shift-and-stack (SAS) processing \citedsp{Document-11013}, because SAS is not a capability being built within the DM system and cannot be done solely by reconfiguring DM pipelines. An example of user-generated pipeline for SAS is described in Section \ref{sec:SPCS}.

\medskip
\noindent \textbf{Stars in the Milky Way and Magellanic Clouds -- } As described in \citedsp{Publication-141}, mini-surveys of the Galactic Plane can better distinguish faints stars from faint red galaxies by including at least 3 filters of coverage (e.g., $izy$; similar to WFD), and could mitigate losses from proper motion and increase the detection rate of stellar flares by obtaining all the images in short time span (i.e., a more concentrated cadence than the WFD).  As described in \citedsp{Publication-145}, applying the nominal DDF observing strategy over the full area of the Large and Small Magellanic Clouds can characterize stellar variability to $M_V<6.5$ on timescales from 15 seconds to 3 days. For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"} \citedsp{Publication-145}. The Twilight survey in Table \ref{tab:ddfms} proposes short exposures to enable bright stars to be put on the same photometric system as the deeper LSST WFD main survey catalog, and enable science that is based on their long monitoring baselines from historical observations. In Chapter 10.4 of \citep{2017arXiv170804058L}, a proposed short-exposure survey of M67 would use the camera's stretch goal of $0.1$ second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as $6$ magnitudes brighter than the saturation level"}. {\bf Summary:} while some of these science goals can be accomplished with standard visits, MW \& L/SMC science goals are likely to request shorter exposure times, perhaps down to $0.1$ seconds. These science goals are also likely to propose cadence and filter distributions that are significantly different from the WFD main survey. From a processing perspective, the science goals depending on shorter exposures will only be able to be met by reconfiguring the DM pipelines if the short exposures can be shown to successfully be processed (with, e.g., instrument signature removal); the science goals can likely be met with data products in the same format as the Prompt or DR Pipeline (i.e., {\tt Source} and {\tt Object} catalogs, single visits and deep CoAdds). Although it is not mentioned in the above paragraph, the MW \& L/SMC science community is also most likely to require special processing to extract information from saturated stars, which is outside the scope of DM. See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.

\noindent \textbf{Exoplanets -- } As described in Section 3.1.2 of \citep{2008arXiv0805.2366I}, transiting exoplanets could be detected with the nominal DDF plan, which would allow for $1\%$ variability to be detected over hour-long timescales; a DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence. \citep{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane, and how microlensing candidates can be found with $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (since microlensing events are slower; these would then require follow-up with external facilities). Dealing with the more crowded fields would be mitigated by the shallower images, in this case. One of the main points of \citep{2013arXiv1304.3455G} is that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped. \textbf{Summary.} Some of these science goals appear possible with standard visit images, and some might request shorter exposures to avoid confusion in crowded fields when the science can be done with brighter stars. From a processing perspective, the science goals are likely to be achievable with reconfigured DM pipelines, but this depends heavily on performance in crowded fields. See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.

\noindent \textbf{Supernovae -- } The nominal DDF plan described in \citep{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses. The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$ \citedsp{Publication-144}. High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \citep{2014ApJ...794...23D}. Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template. This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days. These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss. SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science. The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \citedsp{Publication-142}. Short-exposure observations of bright, nearby SNe may also be useful to include near-peak photometry in the LSST magnitude system, and enable full light-curve analyses. \textbf{Summary.} All of these science goals appear possible with standard visit images (with the exception of a target-of-opportunity short-exposure program to observe bright SNe). From a processing perspecitve, the science goals appear to be accessible with reconfigured DM pipelines to stack and difference the data. In particular, the DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images. See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.

\noindent \textbf{Galaxies -- } The additional depth of a DDF may provide access to a larger collection of low-$\mu$ objects. \citedsp{Publication-142} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies". The DDF stacks could also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. Also, the DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citedsp{Publication-142,Publication-143}. \textbf{Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources. In addition, it seems likely that user-generated algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, and these are beyond the scope of DM.

\noindent \textbf{Weak Lensing -- } The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch). \textbf{Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
%$\bullet$ \textit{Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (\texttt{\Large{Community}} forum) \\
%$\bullet$ \textit{Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline. We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (\texttt{\Large{Community}} forum)  \\
%$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citedsp{Document-10963} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\

>>>>

\setDocAbstract{
The core LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, which is expected to be accomplished with 85--90\% of the observing time.
The remaining 10--15\% of the time will be spent on Special Programs: additional survey areas and/or observing strategies that are driven by specific science goals which build on, or are beyond, the core science pillars of the LSST.

\medskip
This document provides a summary of the Special Programs being considered, the potential diversity of data they might produce (e.g., shorter or longer exposure times), and the role of Rubin Observatory in reducing, processing, and serving data products from Special Programs.
The latter includes a discussion about the relevant system requirements, the needed capabilities of the LSST Science Pipelines and the Rubin Science Platform, and the cases in which Special Programs data products would be user-generated.

\medskip
The specialized, science-specific aspects of Special Programs processing and analysis that are best left to the science community are also described, and illustrated with case-study examples.

\medskip
The main target audience of this document is Rubin Observatory staff -- the construction-era Data Management (DM) team and the operations-era Data Production (RDP) team -- but members of the science community who are planning to use Special Programs to reach their science goals may also find this document useful.
}

@@

Note that there remains some grey area between what is a mini-survey within the WFD, and what is a WFD-like micro-survey.
For example, the last two on the list above (the northern stripe and the Virgo cluster) could also be (or might be in the future) called WFD mini-surveys, and not Special Programs micro-surveys.

@@

\subsection{Rubin Requirements Related to Special Programs}
% \lsrreq, \ossreq, \dmreq
% \reqparam

A detailed list of the requirements related to Special Programs is provided in Appendix~\ref{sec:docrev}.
The most important requirements are summarized below.

\subsubsection{Data Products}

Rubin Observatory and the LSST system (the observatory and the data management systems) are required to process Special Programs data to produce unique and separate data products ``whenever possible" (LSR-REQ-0121).

The term ``whenever possible" includes cases where the original or reconfigured versions of the LSST Science Pipelines can be run, and excludes cases where the development of new algorithms or the allocation of significant additional computational resources are required (LSR-REQ-0121).

The statement ``to produce unique and separate data products" typically refers to producing the same kinds of data products as will be generated by the Prompt and Data Release pipelines (processed visit images, coadded images, difference images, and catalogs of sources and objects for those images).

It is a requirement that the cumulative size of the Special Programs data products generated by Rubin Observatory be no more than $\sim$10\% the size of the Data Release data products (i.e., proportional to the fraction of survey time spent; LSR-REQ-0121).

It is a requirement that these Special Programs data products be distinct, and joinable with (in other words, they can be federated or cross-matched with) the Prompt and Data Release data products (DMS-REQ-0322).

The derivation of value-added data products, such as HiPS or MOC maps, for Special Programs remains an open question (DMS-REQ-0379, 0383).

\subsubsection{Metadata}

In order to support Special Programs processing, the LSST system is required to store metadata that includes program information for every raw image, such as identifiers for images obtained as part of the Main Survey or a Special Program (DMS-REQ-0068).

It is required that this metadata be sufficient for Special Programs to trigger their own real-time data processing recipes ``whenever possible" (DMS-REQ-0320), and be included in alert packets (DMS-REQ-0274).

\subsubsection{Processing}

It is a requirement that Special Programs processing with the Prompt pipeline (or a reconfigured version of it) is subject to the same timescales and latency constraints of 24 hours for the release of Prompt data products and 1 minute for the transmission of Alert packets (DMS-REQ-0344).

It is also a requirement that Special Programs processing be done on timescales intermediate to the Prompt and Data Release processing, ``whenever possible" and whenever necessary to enable the intended science goals of the Special Program (LSR-REQ-0032).

It is a requirement that the LSST system be able to process non-standard visits with short exposure times as low as 1 second, with a discussion note that such short exposures might have degraded image quality (LSR-REQ-0111).

Processing for Special Programs by Rubin Observatory is expected to use no more than $\sim$10\% of computational and storage capacity of the Rubin data processing cluster (i.e., proportional to the fraction of survey time spent and the size of the Rubin-processed Special Programs data products; Section 6 of the DPDD).

\subsubsection{User Processing}

In cases where the science goals of a Special Program require that new algorithms or software be developed, User-Generated pipelines and data products will be needed.

The 10\% of the total data processing capacity that Rubin Observatory is required to reserve for all User-Generated processing includes that applied by users to Special Programs data -- in other words, there is no additional capacity {\it for users} that will be reserved only for Special Programs data (LSR-REQ-0041).

@@

{\bf Solar System Processing -- } Since Solar System Processing takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
As discussed under "Solar System Objects (SSO)" in Appendix \ref{sec:prevpropsp}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
%There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.

@@

\begin{itemize}
\item discoverability of SP data when browsing an all-sky map
\item to query data by tag of WFD, WFD mini-survey, or Special Program (e.g., DDF field, micro-survey identifier)
\end{itemize}

@@

% % % % % % % % % % % % % % % % % %
\clearpage
\bibliography{local,lsst,refs,books,refs_ads}

% % % % % % % % % % % % % % % % % %
\clearpage
\appendix

\input{ap_docrev.tex}

\input{ap_hardbounds.tex}

\input{ap_procbounds.tex}

\input{ap_prevpropsp.tex}
